RILS-ROLS: Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-10-26 doc/paper.tex
Experimental evaluation: working...

Our proposed algorithm was implemented in Python version 3.x. All experiments were performed on a machine xx with xx GHz and a memory limit of xx GB in single-threaded mode. The maximum computation time allowed for each run was limited to 1h minutes, i.e., 3600 seconds.

The following 15 algorithms are compared: \textsc{AI-Feynman}, \textsc{GOMEA}, \textsc{Afp-Fe}, \textsc{Itea}, \textsc{Afp}, \textsc{Dsr}, \textsc{Operon},  a python package \textsc{gplearn}~\cite{stephens2016genetic}, \textsc{Sbp-Gp}, \textsc{Eplex}, \textsc{Bsr}, \textsc{Feat}, \textsc{Ffx}, \textsc{Mrgp} and our \textsc{Rils}-\textsc{Rols} algorithm.

@@

SRBench is an open-source benchmarking project which merge a large set of diverse benchmark datasets, contemporary SR methods as well as ML methods around a shared model evaluation and the environment for analysis, see~\cite{la2021contemporary}. It brought researchers an easy handled environment to evaluation, compare, and analyse their methods. Among all benchmark sets, we involve two real-world inspired benchmark sets into our computational evaluation: 
\begin{itemize}
	\item \textsc{Feynman} benchmark set, inspired from physics and formulas/models that correspond to various natural laws.  
	It consists of $10^5$  samples, each described in~\cite{udrescu2020ai}. 
	\item \textsc{Strogatz} benchmark set from~\cite{la2016inference}. 
	Each dataset is represented by a 2-state system of first-order, ordinary differential equations. 
	The aim of each problem is to predict rate of change of the subsequent state w.r.t\  the current two states on which it depends. These equations describes various non-linear dynamics and exhibiting chaos.

@@

\subsection{Comparison with other methods on real--world benchark sets}
In this section, we evaluate the performance of \textsc{Rils-Rols} algorithm with 15 other competitors from the literature that solve the problem of Symbolic regression. The results are reported in terms of three numerical Tables~\ref{tab:comp_noise0}--\ref{tab:comp_noise001} on real-world benchmark sets, \textsc{Feynman} and \textsc{Storgatz}, utilizing  different values of noise: 0.0 (no noise), 0.001, and 0.01, respectively. Each table consists of three blocks. The first block reports the algorithm whose performances are measured. The second block reports, in two columns,  percentages of instances where the exact model has been found by respective algorithms on both, \textsc{Feynman} and \textsc{Strogatz}, benchmark sets. The third column provides the overall summary percentage of algorithms' success over all instances. The third block displays the results of respective algorithm in the presence of $R^2$ error, that is when $R^2$ error reached the value of 0.999, when the algorithm gets terminated providing a correct output w.r.t.\ the error. Note that the termination may not necessary be triggered by the error equal to 1. This situation is of interest as it may indicate to the existence of an over-fitting. This block consists of three rows that indicate the percentage of instances where respective algorithm is successfully terminated (fulfilling condition $R^2 \geq 1$) on the both real--world benchmark sets separately, and the overall percentage of success rate (over all considered instances).

Note that when the algorithm terminates with $R^2=1$ and $RMSE=0$, the match of found model and the exact model is proved -- the results are reported in the second block. However, in case when $R^2 \neq 1$, that is $R^2 \geq 0.999$, we manually check these cases comparing the match of obtained model and the exact model, indicating the result. \fxnote{Aco, projeri ovu recenicu, da nisam promasio...}

@@

\begin{itemize}
	\item Concerning the exact results in case noise is not presented, the best performing algorithm is our \textsc{Rils}--\textsc{Rols}, able to find the right model on 57.84\% problems instances of the benchmark set \textsc{Feynman}; the same algorithm was even more impressive in solving problem instances from benchmark set \textsc{Strogatz}, begin successful in 83.57\% cases. The second best performing  approach is \textsc{AI-Feynman}, successful in solving 55.78\% and 27.14\% instances of the benchmark set \textsc{Feynman} and \textsc{Strogatz}, respectively. All other approaches are performing significantly worse, and none of them is able to solve more than 30\% instances of any of the two considered benchmark sets. 
	\item   Concerning the exact results when noise is presented with a value of 0.001, our \textsc{Rils}--\textsc{Rols} is still performing reasonably well, having found an exact model in more than 42\%, considering problems from the both benchmark set. Interestingly, \textsc{AI-Feynman} is quite sensible w.r.t.\ the change in noise, as its performances rapidly drop off -- it is successful on just 2.81\% of the all problem instances. Better from \textsc{AI-Feynman} are in this case \textsc{Dsr}, \textsc{Gplearn}, and \textsc{Afp-Fe} in this order, whose percentages of success range from about 15--17\%. 
	\item  Concerning the exact results when noise is high, that is with a value of 0.01,  our \textsc{Rils}--\textsc{Rols} is still reasonably good, having found an exact model for  34.77\% of the all problem instances. The second best is \textsc{Gplearn} which percentage of success is 12.75\%, followed by \textsc{Dsr} which is successful on just 9.23\% of the all problem instances. 
	\item \fxnote{Marko: opisati rezultate za $R^2 >0.999$}
\end{itemize}

>>>>

Our \textsc{Rils}--\textsc{Rols} algorithm is implemented in Python version 3.x. All experiments are performed on a machine xx with xx GHz and a memory limit of xx GB in single-threaded mode. The maximum computation time allowed for each run was set to 1h minutes, i.e., 3600 seconds.

The following 15 algorithms are compared: \textsc{AI-Feynman}, \textsc{GOMEA}, \textsc{Afp-Fe}, \textsc{Itea}, \textsc{Afp}, \textsc{Dsr}, \textsc{Operon}, the \textsc{Gplearn} from python package \textsc{gplearn}~\cite{stephens2016genetic}, \textsc{Sbp-Gp}, \textsc{Eplex}, \textsc{Bsr}, \textsc{Feat}, \textsc{Ffx}, \textsc{Mrgp} and our \textsc{Rils}-\textsc{Rols} algorithm.

@@

SRBench is an open-source benchmarking project which merge a large set of diverse benchmark datasets, contemporary SR methods as well as ML methods around a shared model evaluation and the environment for analysis, see~\cite{la2021contemporary}. It brought researchers an easy handled environment to evaluation, compare, and analyze their methods. Among all benchmark sets, we involve two real-world inspired benchmark sets into our computational evaluation: 
\begin{itemize}
	\item \textsc{Feynman} benchmark set, inspired from physics and formulas/models that correspond to various natural laws.  
	It consists of $10^5$  samples, each described in~\cite{udrescu2020ai}. 
	\item \textsc{Strogatz} benchmark set from~\cite{la2016inference}. 
	Each dataset is represented by a 2-state system of first-order, ordinary differential equations. 
	The aim of each problem is to predict rate of change of the subsequent state w.r.t\  the current two states on which it depends. These equations describes various non-linear dynamics and exhibiting chaos.

@@

\subsection{Comparison with other methods on real--world benchark sets}
In this section, we evaluate the performance of \textsc{Rils-Rols} algorithm with 15 other competitors from the literature that solve the problem of Symbolic regression. The results are reported in terms of three numerical Tables~\ref{tab:comp_noise0}--\ref{tab:comp_noise001}, where two real-world benchmark sets, \textsc{Feynman} and \textsc{Storgatz} are considered, utilizing  different values of the noise levels: 0.0 (no noise), 0.001 (a low-level noise), and 0.01 (a high-level noise), respectively. Concerning the added noise, the White Gaussian noise was added to the target ($y$-axis values) as a fraction of the signal RMS value, also applied in~\cite{la2021contemporary}, i.e. for target noise level $\alpha$, we have
$$ y_{noise} = y + \epsilon, \epsilon \sim \mathcal{N}\left(0, \alpha \sqrt{\frac{1}{N} \sum_{y_i^2}}\right). $$

Each table consist of three blocks. The first block reports the algorithm whose performances are reported. The second block reports, in two columns,  percentages of instances where the model which matches to the exact one has been found by respective algorithms on both, \textsc{Feynman} and \textsc{Strogatz}, benchmark sets. The third column of this block provides the overall summary percentage of success over all instances. The third block displays the results of the respective algorithm in the presence of $R^2$ error which can indicate on  accuracy, i.e. in our cases when the algorithm  at least reached the accuracy ($R^2$ score) of 0.999.  Note that the termination of our algorithm may not necessary be triggered by the error equal to 1. These results are of interest as it may point out the existence of an over-fitting. This block consists of three rows that indicate the percentage of instances where respective algorithm is able to reach the desired accuracy (fulfilling condition $R^2 \geq 0.999$ before a termination due exceeding the time limit) on the both real--world benchmark sets separately, and the overall percentage of success rate (over all considered instances) w.r.t. the same criterion of success.

%Note that when the algorithm terminates with $R^2=1$ and $RMSE=0$, the obtained model and the exact model are always match (but also checked manually). However, in case when the algorithms deliver an error $R^2 \geq 0.999$, but $R^2 \neq 1$, these cases are all  manually checked by comparing weather the produced model match the exact model.

In short, the exact results, thus the number of instances (in percentages) for which the output of respective algorithm match to the exact model, are displayed in the second block of each of the tables. The numbers of instances (in percentages) for which each algorithms deliver a precise enough model (i.e. satisfying $R^2\geq 0.999$ upon its termination) are displayed in the third block of each of the tables.

@@

\begin{itemize}
	\item Concerning the exact results in case no noise is utilized in the instance problems, the best performing algorithm is our \textsc{Rils}--\textsc{Rols}, able to find the right model on 57.84\% problems instances of the benchmark set \textsc{Feynman}; the same algorithm was even more impressive in solving problem instances from benchmark set \textsc{Strogatz}, begin successful in 83.57\% cases. The second best performing  approach is \textsc{AI-Feynman}, successful in solving 55.78\% and just 27.14\% of all instances of the benchmark set \textsc{Feynman} and \textsc{Strogatz}, respectively. All other approaches are performing significantly worse, and none of them is able to solve more than 30\% instances of any of the two considered benchmark sets. 
	\item   Concerning the exact results when the target noise is presented with a level of 0.001, our \textsc{Rils}--\textsc{Rols} is still performing reasonably well, having found an exact model in more than 42\% cases considering the problems from the both benchmark sets. Interestingly, \textsc{AI-Feynman} is quite sensible w.r.t.\ the increase in the target noise level, as its performances rapidly drop down -- it is successful on just 2.81\% of the all problem instances. Better from \textsc{AI-Feynman} are in this case the \textsc{Dsr}, \textsc{Gplearn}, and \textsc{Afp-Fe} in this order of success, whose percentages of success rate range from $\approx$15--17\%. 
	\item  Concerning the exact results when the presence of noise is significant, that is with a level of 0.01,  our \textsc{Rils}--\textsc{Rols} is still performing reasonably well, having found an exact model for  34.77\% of the all problem instances. The second best is \textsc{Gplearn} which percentage of success is just 12.75\%, followed by \textsc{Dsr} which is successful for just 9.23\% of the all problem instances.

	\item When comparing performances of the algorithms in terms of number of instances for which the algorithm's termination led to an accurate model, that is $R^2 \geq 0.999$, in case when no noise is integrated, our \textsc{Rils}-\textsc{Rols} is successful with the rate of success of 83.38\%. However, in contrast with the exact model percentages comparisons from above, there are a few methods reaching even better percentages of success than ours: \textsc{Mrgp}, and \textsc{Operon}, with 92.69\%, and 86.93\%. On the other hand, as the  exact model percentage rates of these two approaches are rather low (0\% and 16\%, respectively) these indicate the presence of a high level over-fitting  of the later two algorithms. Thus, they are able to obtain a model that is accurate enough on the training data, but most of the produced models are in essence wrong.

   \item  When comparing the performances of the competitors in terms of number of instances for which the  termination led to an accurate enough model, in the presence of the noise level of 0.001, our \textsc{Rils}--\textsc{Rols} method has 78.62\% of a success rate, concerning all considered instances.  The \textsc{Operon} solver only has a better percentage of success among the competitors, which is 83.08\%. However, the over-fitting is conspicuous as this algorithm is rarely able to obtain the model which matches to the exact one (which is successful for 0.31\% cases).

   \item   When comparing the performances of the competitors in terms of number of instances for which the  termination led to an accurate enough model, in the presence of the high--level noise of 0.01, our \textsc{Rils}--\textsc{Rols} method is still highly accurate, delivering  62.92\% of a success rate. Interestingly, it is the best performing in terms of the percentage rate among other competitors. Just for the information, the second best approach is \textsc{Dsr}, with a 16.92\% success rate. It can be concluded that our \textsc{Rils}--\textsc{Rols} method does not pay a big price in presence of a higher noise, like the other methods do, and seem to be very sensible w.r.t. the noise increase. Our \textsc{Rils}--\textsc{Rols} method still delivers high-accuracy models for the noisy input data, but also many of them match to the exact solutions (models).

\end{itemize}
